Analysis of gene regulatory properties underlying trait pleiotropy

### 2022-10-17 presentation_gold2022_paris/presentation_gold2022_paris.tex
ENH
ENH presentation o

\begin{itemize}
\item Genetic loci with higher frequency in individuals with a given disease
\item A very active field with many studies
\item Many genetic variants are involved in several phenotypes
\end{itemize}
%
\vfill
%
Limitations
%
\vfill
%
\begin{itemize}
\item Correlation between variants, ie. LD
\item Which cell types are causal to the disease?
\item Over 90\% GWAS variants fall in non-coding regions
\end{itemize}
%
\vfill
%
These limitations can be partially addressed by looking at eQTLs in different tissues

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Integration of GWAS and eQTL variants:  Colocalisation analysis}

\begin{figure}[!]
\includegraphics[width=0.6\textwidth]{/home/gonzalez/Repositories/gwas2eqtl_pleiotropy/presentation_gold2022_paris/fig/doi_10.3389_fgene.2020.00424_fig4.png}
\end{figure}

@@

\begin{itemize}
\item To carry out a large scale colocalization analysis between GWAS and eQTL variants
\item To identify pleiotropic genetic loci
\item To characterize the regulatory causas of this pleiotropy
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item IEA OpenGWAS database
\item EBI eQTLs database
\item CRAN - Package coloc
\end{itemize}

\let\thefootnote\relax\footnotetext{https://gwas.mrcieu.ac.uk}
\end{frame}

@@

\begin{itemize}
\item Colocalization of variants between 418 GWAS and 127 eQTL studies, ...
\end{itemize}

@@

\begin{itemize}
\item Distance between GWAS traits based on eQTL beta
\item Clustering of immune diseases coherent
\item eQTL beta of GWAS variants coherent
\end{itemize}

\begin{figure}[!]
\includegraphics[width=0.6\textwidth]{/home/gonzalez/Repositories/gwas2eqtl_pleiotropy/out/gwas418/pval_5e-08/r2_0.1/kb_1000/window_1000000/plthtmp_disease_comorbidity_matrix.py/corr.png}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Where are the pleiotropic variants?}

    \begin{center}
\includegraphics[width=0.8\textwidth]{/home/gonzalez/Repositories/gwas2eqtl_pleiotropy/out/gwas418/pval_5e-08/r2_0.1/kb_1000/window_1000000/pltbar_vep_consequence.py/vep.png}
     \end{center}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Do pleiotropic variants bind more transcription factors?}

@@

@@

Characteristics of pleiotropic variants
%
\begin{itemize}
\item Variant pleiotropy is correlated in regions
\item Enrichment of variants upstream (incl. 5'UTR) and downstream of genes (incl. 3'UTR)
\item Enrichment of transcription factor and CRM binding
\item Increased number of associated egenes and etissues
\end{itemize}
%
\vfill
%
Other outcomes of this study
%
\begin{itemize}
\item A large analysis of GWAS/eQTL colocalisations
\item A list of pleiotropic variants and regions
\item Analysis of regulatory properties of pleiotropic variants
\end{itemize}

>>>>

\begin{itemize}
\item Genetic loci with higher frequency in individuals with a given disease
\item A very active field with many studies
\item Many genetic variants are involved in several phenotypes
\end{itemize}
%
\vfill
%
Limitations
%
\begin{itemize}
\item Correlation between variants, ie. LD
\item Which cell types are causal to the disease?
\item Over 90\% GWAS variants fall in non-coding regions
\end{itemize}
%
\vfill
%
These limitations can be partially addressed by looking at eQTLs in different tissues

\let\thefootnote\relax\footnotetext{Cano-Gamez et al. 2020. doi:10.3389/fgene.2020.00424}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Expression quantitative trait loci (eQTL)}

\begin{columns}
\begin{column}{0.5\textwidth}
    \begin{center}
\includegraphics[width=\textwidth]{/home/gonzalez/Repositories/gwas2eqtl_pleiotropy/presentation_gold2022_paris/fig/doi_10.3389_fgene.2020.00424_fig4a.jpg}
     \end{center}
\end{column}
\begin{column}{0.5\textwidth}

\end{column}
\end{columns}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Colocalisation analysis of GWAS and eQTL variants}

\begin{columns}
\begin{column}{0.5\textwidth}
    \begin{center}
\includegraphics[width=\textwidth]{/home/gonzalez/Repositories/gwas2eqtl_pleiotropy/presentation_gold2022_paris/fig/doi_10.3389_fgene.2020.00424_fig4a.jpg}
     \end{center}
\end{column}
\begin{column}{0.5\textwidth}
    \begin{center}
\includegraphics[width=\textwidth]{/home/gonzalez/Repositories/gwas2eqtl_pleiotropy/presentation_gold2022_paris/fig/doi_10.3389_fgene.2020.00424_fig4bc.png}
     \end{center}
\end{column}
\end{columns}

@@

\begin{enumerate}
\item Large scale colocalization analysis between GWAS and eQTL variants
\item To identify pleiotropic genetic loci
\item To characterize the regulatory causes of trait pleiotropy
\end{enumerate}
\end{frame}

@@

\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item IEA OpenGWAS database$^1$: 244,881e6 associations from 42e3 GWAS datasets
\item EBI eQTLs database$^2$: Processed gene expression and splicing QTLs from all public studies on human.
\item CRAN - Package coloc$^3$: Genetic colocalisation analysis of two phenotypes to look for shared common genetic causal variants
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\begin{center}
\includegraphics[width=\textwidth]{/home/gonzalez/Repositories/gwas2eqtl_pleiotropy/presentation_gold2022_paris/fig/pgen.1004383.g001.png}
\end{center}
\end{column}
\end{columns}

\let\thefootnote\relax\footnotetext{$^1$https://gwas.mrcieu.ac.uk}
\let\thefootnote\relax\footnotetext{$^2$https://www.ebi.ac.uk/eqtl}
\let\thefootnote\relax\footnotetext{$^3$https://cran.r-project.org/web/packages/coloc/index.html}
\end{frame}

@@

\begin{itemize}
\item Colocalization analysis of 418 GWAS and 127 eQTL studies (53e3 pairs)
\item 143e3 shared causal variants with probability $\geq$ 0.8
\item Percentage of explained GWAS loci: 0$\%$ - 77$\%$ (Average 33$\%$)
\end{itemize}

@@

Distance between GWAS traits based on eQTL beta

\begin{figure}[!]
\includegraphics[width=0.5\textwidth]{/home/gonzalez/Repositories/gwas2eqtl_pleiotropy/out/gwas418/pval_5e-08/r2_0.1/kb_1000/window_1000000/plthtmp_disease_comorbidity_matrix.py/corr.png}
\end{figure}

Clustering coherent

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Where are the pleiotropic variants?$^1$}

    \begin{center}
\includegraphics[width=0.8\textwidth]{/home/gonzalez/Repositories/gwas2eqtl_pleiotropy/out/gwas418/pval_5e-08/r2_0.1/kb_1000/window_1000000/pltbar_vep_consequence.py/vep.png}
     \end{center}
varianb
\let\thefootnote\relax\footnotetext{$^1$https://www.ensembl.org/vep}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Do pleiotropic variants bind more transcription factors?$^1$}

@@

\let\thefootnote\relax\footnotetext{$^1$https://remap.univ-amu.fr}

@@

Regulatory characteristics of pleiotropic variants
%
\begin{itemize}
\item Variant pleiotropy is correlated in regions
\item Enrichment of variants upstream (incl. 5'UTR) and downstream of genes (incl. 3'UTR)
\item Enrichment of transcription factor and CRM binding
\item Increased number of associated egenes and etissues
\end{itemize}
%
\vfill
%
Other outcomes of this study
%
\begin{itemize}
\item A large analysis of GWAS/eQTL colocalisations
\item A list of pleiotropic variants and regions
\end{itemize}
